\lab{Applications}{Latent Semantic Indexing}{Latent Semantic Indexing}

\objective{Understand the basics of latent semantic indexing}

Suppose we have a large collection of documents, one of which is an article known by us to be about PCA. We think we understand PCA, but we would like to find a different article that provides the details of PCA from a slightly different perspective (maybe written by a different author). How do we find this article from our entire collection of documents? We might consider simply choosing the article that contains the acronym \emph{PCA} the greatest number of times, though it is possible that it is an article about a specific data set, in which they frequently refer to the results obtained from PCA without actually explaining how the method works.

A better way is to use a form of PCA itself on the collection of the documents! The idea is that we define the unique set of words occuring in the entire collection of documents as a vocabulary, and represent each document as a vector of count sums. In this case, the $i^{th}$ entry of a document's vector is $t\!f_{i}$, where $t\!f_{i}$ is defined to be the number of times term $i$ occurs in the document. We then represent the entire collection of documents as an $m \times n$ matrix $t\!f$, where $m$ is the length of the vocabulary and $n$ is the number of documents in our collection, each column being a document vector. As expected, we let $t\!f_{ij}$ be the number of times term $i$ occurs in document $j$.

We perform PCA on $t\!f$ without centering the data so that we may retain the sparsity of $tf$. We now have $t\!f = T\Sigma D^{H}$. The $i^{th}$ row of $D$ is a vector corresponding to the $i^{th}$ document. We typically truncate these vectors, similar to identifying the important principal components of $t\!f$. We let $D_{i}$ denote the $i^{th}$ truncated row of $D$. Given a document $i$, we would like to find the document $j$ most similar to $i$, given that $i \neq j$. What we really need to compare are the vectors $D_{i}\Sigma$ and $D_{j}\Sigma$. Our metric of choice is the angle between two vectors, which is easily computable by considering the identity $$\cos \theta_{ij} = \frac{D_{i}\Sigma \cdot D_{j}\Sigma}{\|D_{i}\Sigma\| \|D_{j}\Sigma\|}$$ where $\theta_{ij}$ is the angle between $D_{i}\Sigma$ and $D_{j}\Sigma$. If two document vectors have a small angle between them, then they are similar; a large angle signifies that they are quite different. Since $\cos \theta_{ij}$ is large whenever $\theta_{ij}$ is small, we seek to find $$\argmax_{j} \frac{D_{i}\Sigma \cdot D_{j}\Sigma}{\|D_{i}\Sigma\| \|D_{j}\Sigma\|}.$$

While this is the idea, in practice there are a few other steps we must take prior to the matrix decomposition. In any language, some words are so common that they provide no important information. We call these \emph{stop words}. Examples in English include \emph{the, a, an, and, I, we, you, it, there}, etc. Before processing any set of documents, we must remove all occurrences of stop words. We are going to consider the collection of all State of the Union addresses from 1945 to 2013, but these speeches are in an uncleaned format in the folder \emph{Addresses}, so we must first process them.

\begin{problem}
Using the function \emph{loadStopwords} and the file \emph{stopwords.txt} provided, read in the stop word list in Python. Use the functions \emph{termlist}, \emph{removeStopwords}, \emph{uniq}, and \emph{parseDocument} provided to create an array containing all of the unique terms encountered in the corpus that are not included in the stop word list. This is called the \emph{term list}.
\end{problem}

\begin{problem}
Write a function that accepts a file name, a list of stop words, and the term list found above, that will return a vector whose $i^{th}$ entry is the number of times the $i^{th}$ word from the term list occurs in the .txt file. Note that no word in the stop word list should ever contribute to any entry of this vector.
\end{problem}

\begin{problem}
Write a function that accepts a directory, a list of stop words, and the term list, that will make the term frequency matrix $t\!f$ as described above. Each column in $t\!f$ should be the value returned by the function written in the previous problem for a particular file. In particular, $t\!f$ should be a $m \times n$ matrix, where $m$ is the length of the term list and $n$ is the number of .txt files in the given directory.
\end{problem}

Not only is removing a predetermined set of stop words a good idea, but it's also important to consider that words appearing in few documents tend to provide more information than words occurring in every document. For example, while the word \emph{war} might not be considered a stop word, it is likely to appear in quite a few addresses, while \emph{Afghanistan} will not. Thus two speeches sharing the word \emph{Afghanistan} ought to be considered more related than two speeches sharing the word \emph{war}. So while $t\!f_{ij}$ is a good measure of the importance of term $i$ in document $j$, we also need to consider some kind of global weight for each term $i$, indicating how important the term is over the entire collection. There are a number of different weights we could choose, but we choose to employ the following particular approach.

Let $g\!f_{i}$ be the total number of times term $i$ appears in the entire collection of documents. Define $$p_{ij} = \frac{t\!f_{ij}}{g\!f_{i}}.$$ We then let $$g_{i} = 1 + \sum_{j=1}^{n} \frac{p_{ij} \log (p_{ij} + 1)}{\log n}$$ where $n$ is the number of documents in the collection. We call $g_{i}$ the global weight of term $i$. We replace each term frequency in the matrix $tf$ by weighting it globally. Specifically, we define $$a_{ij} = g_{i} \log (t\!f_{ij} + 1),$$ so we can consider these values to be a matrix $A$, which will take the place of $tf$ as a matrix whose entries are both locally and globally weighted.

\begin{problem}
Write a function that accepts the term frequency matrix returned by your previous function, and which returns the matrix $A$ as described above. Perform PCA on $A$ as described above, remembering to \emph{not} center the data. Truncate the document vectors to retain the first $80\%$ of the variance in the data.
\end{problem}

\begin{problem}
Write a function that accepts the truncated document matrix, the list of file names for the documents, and an index $i$, and which returns the file name of the document most similar to the $i^{th}$ document. Find the speeches most similar to each of President George W. Bush's speeches, as well as the speeches most similar to each of President Obama's speeches. What conclusions about their speeches can you draw?
\end{problem}

