\lab{Algorithms}{Np-Hard Simplex}{Np-Hard Simplex}
\label{lab:Simplex}
\objective{In this lab, you will see the downside of the Simplex method.}

The simplex algorithm is one of the most used in business.
The Computing in Science and Engineering journal listed Simplex as one of the top ten algorithms of the twentieth century.
Despite its popularity, like any other algorithm , simplex has drawbacks.  

The Victor Klee and George Minty Cube wrote a paper in 1972 called, "How Good is the Simplex Algorithm?" 
In their paper, they give several examples of polytopes that struggle with the Simplex algorithm. 


\section*{Example}

Consider the following linear program from Klee-Minty.

\begin{align*}
\text{max } & 2^{n-1}x_1 & + & 2^{n-2}x_2  & + & \cdots & + & 2x_{n-1} & + & x_n\\                     
\text{subject to } & x_1 &  &  &  &  &  &  &  &\leq 5\\
& 4x_1 & + & x_2 &  &  &  &  &  &\leq 25\\
& 8x_1 & + & 4x_2 & + & x_3 &  &  &  &\leq 125\\
& \vdots & &      &   &     &  &  &  &\vdots\\ 
& 2^n x_1 & + & 2^{n-1} x_2 & + & \cdots & + & 4x_{n-1} & + x_n &\leq 5\\
\end{align*}

When $n = 3$, we have the initial tableau
$$
\begin{bmatrix}
0 & 4 & 2 & 1 & 0 & 0 & 0\\
0 & 1 & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 1 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 1 & 0 & 0 & 0\\
5 & -1 & 0 & 0 & 0 & 0 & 0\\
25 & -4 & -1 & 0 & 0 & 0 & 0\\
125 & -8 & -4 & -1 & 0 & 0 & 0\\
\end{bmatrix}
$$. 

After the first pivot with $x_1$ leaving and $s_1$ entering, we have the tableau

$$
\begin{bmatrix}
20 & 0 & 2 & 1 & -4 & 0 & 0\\
5 & 0 & 0 & 0 & -1 & 0 & 0\\
0 & 0 & 1 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 1 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & 1 & 0 & 0\\
5 & 0 & -1 & 0 & 4 & 0 & 0\\
85 & 0 & -4 & -1 & 8 & 0 & 0\\
\end{bmatrix}
$$. 

\textbf{Problem 1}

What is the final tableau for this Klee-Minty example with $n=3$?
How many iterations does it take to arrive at the final tableau?

\textbf{Problem 2}
Using problem 1 as a guide, guess the optimum value of the Klee-Minty example with $n=20$.
Then find the maximum value for the Klee-Minty example with $n=20$. 
How many iterations does it take to arrive at the final tableau?
How long does the program take to run for only $20$ variables?

Klee and Minty show that for this example, the worst case has exponential time complexity.
With only $n$ constraints and $n$ variables, the simplex algorithm goes through $2^n$ iterations.
This is because there are $2^n$ extreme points and when starting at the point $x=0$, the simplex algorithm goes through all of the extreme points before reaching the optimal point $(0,0,\dots, 0, 5^n)$.
Other algorithms, such as the interior point method, solve this problem much faster because they are not constrained to follow the edges. 