\lab{Algorithm}{Kruskal's and Prim's Algorithm}{Kruskal's and Prim's Algorithm}
\label{Ch:Kruskal}

\objective{This section teaches about how minimum spanning tree for a connected weighted graph using Kruskal's and Prim's Algorithm.}

\section*{Weight Graphs and Spanning trees}

%Lab \ref{Markov}

Remember that a graph is composed of two sets: a set of nodes and a set of edges that connect these nodes. 

A graph is directed if connections are uni-directional, and undirected if they are bi-directional. The above graphic shows an undirected graph. A weighted graph is a graph where each edge has a value associated with it. A connected graph is there is a path, or a set of edges, that connect every two nodes together. We can write a matrix that describes this type of graph. We let each row of our matrix represent our starting point and each column represent our destination. We put the value of the edge if there is a path and a 0 if there is not. For the above graph we generate the following matrix:

\[
A = \begin{pmatrix}
0 & 1 & 0 & 0 & 1 & 0\\
1 & 0 & 1 & 0 & 1 & 0\\
0 & 1 & 0 & 1 & 0 & 0\\
0 & 0 & 1 & 0 & 1 & 1\\
1 & 1 & 0 & 1 & 0 & 0\\
0 & 0 & 0 & 1 & 0 & 0
\end{pmatrix}
\]

This matrix is called an adjacency matrix. Note that this matrix is symmetric, since the graph is undirected. 

Another way to store the graph is as a list of the edges and their weights that are connected. it would look like.

$[(   ),
 (   ),
 (   ),
 (   ),
 (   ),
 (   ),
 (   ),
 (   ),
 (   ),]$

For this lab, we will be forcusing on undirecting weighted graphs. 

A spanning tree of a connected graph G is is an undirected graph that contains all the nodes of G, a subset of the edges, and contains no cycles. A cycle, for undirected graphs, is a path where you start and and on the same node without using an edge more than once. The minimum spanning tree of a weighted, undirected graph is the spanning tree where the sum of the weights of the subset of edges is less than or equal to the sum of the weights of every other spanning tree. Both Kruskal's and Prim's Algorithm for the minimal spanning tree of a weighted directed graph.

\section*{Kruskal's algorithm}

Given a weighted directed graph G with $n$ nodes, Kruskal's algorithm finds the minimum by sorted the edges from smallest to largest. Starting from the smallest it adds the edges in order to a tree as long as the addition of a new edge does not create a cycle. When the $n-1$ edges have been added the algorithm stops. 

\begin{problem}
Write Kruskal's algorithm. Test your algorithm on random symmetric matrices. Try sparse symmetric matrices (though remember that the matrix needs to represent a connected graph). use the data from MSTdata.txt to test your tree. open the file by import pickle then pickle.load(open("MSTdata.txt")).
\end{problem}

\section*{Prim's algorithm}

Prim's algorithm accomplishes the same thing as Kruskal's algorithm, but with a different approach. Starting from any node it adds the edge with the least weight. Those two nodes form a tree. It than adds edges with the least weight to any node in the tree to any node not in the tree until all the nodes have been added. 

\begin{problem}
Write Prim's algorithm. Test your algorithm on random symmetric matrices. Then compare the speed of Prim's algorithm with Kruskal's algorithm. Use the same data as the previous problem. 
\end{problem}
