\lab{Algorithms}{Recursion and Persistent Variables}{Recursion and Persistent Variables}

\objective{We will briefly discuss an important technique in programming known as recursion. Then we will discuss the use of persistent variables in MATLAB.}

A recursive function is one that calls itself. For example we could write a factorial function as follows:

\begin{lstlisting}[style=matlab]
function out = myFactorial(n)
if n ~= 1
     out = n*factorial(n-1);
else
     out = 1;
end
\end{lstlisting}

This function calculates factorial recursively. It calls \li{factorial(n-1)} until $n = 1$, at which point each function finally terminates.

One critical aspect of this code is the fact that if n == 1, out = 1. This is known as the base case. The base case is the case where the recursion breaks down, or in other words where the function stops calling itself. Every recursive algorithm must have some sort of base case, or else it will call itself indefinitely.

In fact this function only works on positive integers. On any other input the function never terminates. Thus we should modify the function as follows to avoid problems:

\begin{lstlisting}[style=matlab]
function out = myFactorial(n)
if (n <= 0 || rem(1,n) ~= 0)
	error('input must be a positive integer');
end
	
if n ~= 1
     out = n*factorial(n-1);
else
     out = 1;
end
\end{lstlisting}

This practice (of writing code to handle bad user input) is known as defensive programming. It is good to get into this habit.

Here's a simple combinatoric example. Suppose our goal is to enumerate every possible way to select six numbers out of a list of ten (note that we are talking combinations not permutations). We could use some sort of complicated \li{for} loop structure like this:

\begin{lstlisting}[style=matlab]
for i = ...
    for j = ...
        for k = ...
            for l = ...
            .
            .
            .
\end{lstlisting}

However, we wish to stress emphatically, that this is not the right way to do any problem ever. Once you write more than three for loops you are probably skirting on the edges of intractability (in fact, any time you attack a problem that suggests this approach you are likely attacking a problem that is inherently intractable). However, oftentimes such a problem can at least be programmed using recursion.

For example in our combinatorics problem above we could write something like the following:

\begin{lstlisting}[style=matlab]
function out = Combinations(values,k)
%This functions outputs all possible combinations of k elements from the
%vector values.
if min(size(values)) ~= 1
    error('values must be a vector');
end
if k > length(values)
    error('k must be smaller than length(values)');
end
if (k <=0 || mod(k,1) ~= 0) 
    error('k must be a positive integer'); 
end

%Make input vectors column vectors
if size(values,2) > size(values,1)
    values = values';
end

out = []; 
n = length(values); 
if k == 1
    out = values; 
else
    
%This loop iterates through all of elements of the vector values that have at least k
%elements after them (inclusive). For each element it then calls
%Combinations(values(i+1:end),k-1), which returns combinations of size k-1
%for the elements succeeding the current element. This is so that we do not
%get repeats of combinations.
for i = 1:n-(k-1)
    % Calculate the number of possible combinations (to allow proper
    % concatenation in the recursive call.
    numCombs = factorial(n-i)/((factorial(k-1))*(factorial(n-i-(k-1))));
    %This is the recursive call.
    out = [out;[values(i)*ones(numCombs,1), Combinations(values(i+1:end),k-1)]];
end
end
\end{lstlisting}

This solution utilizes recursion to make the coding much more simple. Write this code up and understand why it does what it does, and how the recursion works.

One note of importance: just because a recursive algorithm exists does not mean that it is the correct way to solve the problem. The factorial example makes this readily evident. Generally speaking recursion slows things down, so it should be avoided when possible. However, it can provide elegant solutions to difficult problems, and is a good arrow to have in your quiver.

\begin{problem}
Use recursion to calculate the determinant of a matrix using cofactor expansions (this is known as Laplace's formula). This can be done in four lines. You can see how simple recursive programs can be. Now time your function and compare its performance to finding the determinant using the LU decomposition that you wrote earlier. How does it fare? You will probably notice that the recursive method is much slower for even moderate values of n. Why is this? Laplace's formula is $O(n!)$, so even though it's very easy to code it's terribly ineffective in general.
\end{problem}

\section*{Persistent Variables}

One of the crucial concepts in computer programming is known as scoping. A variable's scope is the function that it lives in. When we write functions the only variables that we have to worry about are the ones that are passed into it. This is because my function has its own scope. A variable that only lives in a specific function (as most variables should) is known as a local variable.

Persistent variables are a tool that MATLAB uses to ease program complexity. A local variable that is declared to be persistent will have the same value the next time the parent function is called. We will illustrate using a function that calculates the fibonacci numbers. In this example we will also practice using recursive functions.

Write the following function:

\begin{lstlisting}[style=matlab]
function out = fib1(a,b,n)
%% calculate the nth fibonacci number using the seeds a,b
if n ==1
	out = a;
else if n == 2
        out = b;
else
        out = fib1(a,b,n-1) + fib1(a,b,n-2);
end
\end{lstlisting}

This function is very easy to read, and directly applies the recursive definition of the fibonacci sequence. However, there is a problem. Consider the function calls made for $n=5, a = 1, b = 1$:

\begin{align*}
fib(5) &= fib(4) + fib(3) = (fib(3) + fib(2)) + (fib(2) + fib(1)) \\
       &= ((fib(2) + fib(1)) + fib(2)) + (fib(2) + fib(1))
\end{align*}

It is clear that many calculations are being repeated unnecessarily. One possible way to solve this problem is to use a \li{for} loop instead. Try writing the following function:

\begin{lstlisting}[style=matlab]
function out = fib2(a,b,n)
%% calculate the nth fibonacci number using the seeds a,b
if n == 1
	out = a;
else if n == 2
        out = b;
else
        x = [a b];
        for i = 1:n-2
            x = [b sum(x)];
        end
        out = x(2);
end
\end{lstlisting}

This version has the advantage that there are no repeated calculations. However, suppose that we are expecting a and b to usually remain the same. Each time \li{fib} was called all of the calculations would have to be redone. This is where persistent variable can be useful. If a variable is declared persistent its value will be stored for the next time the function is called. In this example we can store the entire fibonacci sequence as a persistent variable and only recalculate as necessary. The following code implements this approach:

\begin{lstlisting}[style=matlab]
function out = fib3(a,b,n)
%% calculate the nth fibonacci number using the seeds a,b
persistent f
if length(f) < 2 || f(1) ~= a || f(2) ~= b
    f = [a b];
end
for k = (length(f) + 1):n
    f(k) = f(k-2) + f(k-1);
end
out = f(n);
\end{lstlisting}

Test this code out. Try testing its performance against the other versions (you could try calculating the same fibonacci number several thousand times). You should see a significant difference.

We finish this section with a word of caution. Persistent variables require significant overhead, and accessing them is much slower than accessing normal variables. They should accordingly be used sparingly.

\begin{problem}
Use persistent variables to speed up the calculations of the \li{expss} function that you wrote in chapter one (hint: certain operations, such as factorial, are costly). Test your new function's performance against the old one. How does it perform?
\end{problem}

\begin{problem}
Compare the performance of the \li{fib3} versus \li{fib2}, when the seeds are random every time. You should run it several thousand times in your test. What do you notice? \li{fib3} is much slower because accessing a persistent variable is very costly. This reminds us that using persistent variables should be avoided in most situations.
\end{problem}
