\lab{Applications}{Orthogonal Polynomials}{Orthogonal Polynomials}
\label{OrthoPoly}

\objective{Teach about orthogonal functions, and applications of function spaces.}

Up to this point, our discussion of inner product spaces has primarily focused on $\mathbb{R}^{n}$ and $\mathbb{C}^{n}$. In these cases the power of the inner product is generally already encoded in the structure that we study in elementary linear algebra. However, inner products are much more useful in the context of infinite vector spaces, specifically in studying spaces of functions.

Consider the following inner product:

\[
<f,g> = \int_{-1}^1 f(x) g(x) w(x) dx
\]

This is known as a weighted inner product (w(x) is known as a weighting function). This inner product is over the space $L^2([-1,1])$. We can consider other intervals of integration as desired.

Recall that we call two vectors orthogonal if $<x,y> = 0$. Using this definition we can find functions that are orthogonal.

\begin{example}
The sequence of functions
\[
1,sin(x),cos(x),sin(2x),cos(2x)
\]
is an orthogonal set on $L^2([-\pi,\pi])$, when $w(x) = 1$. This is clear by integrating using the following identities:
\begin{align*}
2 cos(nx)cos(mx) &= cos(n+m)x + cos(n-m)x \\
2 sin(nx)sin(mx) &= cos(n-m)x - cos(n+m)x \\
2 cos(nx)sin(mx) &= sin(n+m)x - sin(n-m)x
\end{align*}

The fact that these functions are orthogonal is crucial in proving that we can approximate any continuous function arbitrarily well using the fourier transform.
\end{example}

\begin{problem}
The gram-schmidt process can be used to transform a linearly independent set $y_k$ into an orthonormal set $x_k$. The process is defined inductively by:

\[
w_1 = y_1
w_k = y_k - \sum_{n=1}^{k-1} <y_k,x_n>x_n, x_k = \frac{w_k}{||w_k||}
\]

Using $w(x) = 1$, apply the gram-schmidt process to the set of polynomials $t^n$. Find the terms for $n \leq 4$. This set of orthogonal polynomials is known as the Legendre Polynomials.
\end{problem}
\begin{problem}
Prove that the Chebyshev Polynomials form an orthogonal set when $w(x) = \frac{1}{\sqrt{1-x^2}}$. Hint: Use the cosine definition and use the change of variables $x = cos(u)$.
\end{problem}

There are several canonical sets of polynomials that are found using this process. Each is based upon a different weighing function. Some examples include the Hermite polynomials and the Laguerre Polynomials.

One useful application of orthogonal polynomials is that it allows us to easily approximate functions.

\begin{problem}
Find a,b,c that minimize
\[
\int_{-1}^1 |x^3-ax^2-bx-c| dx
\]
Hint: In the framework of calculus this problem would be fairly difficult. However, in the framework of inner product spaces this is much less difficult. We are simply projecting $ax^2-bx-c$ onto $x^3$. You can even reuse the calculations you made in finding the Legendre polynomials.
\end{problem}
