\lab{Algorithms}{Gaussian Quadrature}{Gaussian Quadrature}
\label{Lab:GaussQuad}

Recall from the Newton-Cotes integration lab that certain functons will not be well approximated when we use uniformly spaced points.
In this section we will use unevenly spaced points to mitigate this problem.
The technique we will use is called Guassian Quadrature.
Like our previous integration techniques, we choose a set of points $x_i$ and weights $w_i$ and approximate
\[
\int_{a}^b f(x) dx \approx \sum_{i=1}^n w_if(x_i).
\]

There are some special considerations when we do guassian quadrature.
First, we are required to choose a weight function $W(x)$.
This function determines both the $x_i's$ and the $w_i's$.
Theoretically, the weight function determines which set of orthogonal polynomials we are using to approximate the function we are trying to integrate.

The second consideration is that the weight function also dictates the interval over which the integration will occur.
For example, if we choose the weight function as $W(x) = 1$ over $[-1,1]$, then we may only integrate functions on $[-1,1]$.
We can get around this issue by changing the interval of integration using the following formula
\[
\int_a^b f(x) dx = \frac{b-a}{2}\int_{-1}^1 f(\frac{(b-a)}{2}z + \frac{a+b}{2})dz.
\]
Once we have changed the interval, we may apply quadrature to the integral from $-1$ to $1$ and then scale it appropriately to get the answer we want.
\[
\int_a^b f(x) dx \approx \frac{b - a}{2} \sum_i w_if(\frac{(b-a)}{2}x_i + \frac{(b+a)}{2})
\]

\begin{problem}
Write a function that will accept a function and limits of integration.
Use the change of interval formula and draw a graph of the function on $[-1,1]$.
\end{problem}

We will now give an example with known weights and points.
We will use the constant weight function $W(x) = 1$ from $-1$ to $1$.
This weight functions corresponds to the Legendre polynomials.
We will calculate the integral of $f(x) = sin(x)$ from $-\pi$ to $\pi$ with $5$ interpolation points.

First, we change the interval from $[-\pi, \pi]$ to $[-1,1]$.

\begin{lstlisting}
import numpy as np
a, b = - np.pi, np.pi

#f is the function that we want to integrate
f = np.sin

#g is the function with the interval changed
g = lambda x: f((b - a) / 2 * x + (a + b) / 2 * x)
\end{lstlisting}

The weights and points are given in order in Table \ref{intro_table}.
We put them into an array here.
\begin{center}
\begin{tabular}{|c|c|}
\hline
point $x_i$ & weight $w_i$ \\
\hline
$-\frac{1}{3}\sqrt{5 + 2\sqrt{\frac{10}{7}}}$ &  $\frac{322-13\sqrt{70}}{900}$ \\
\hline
$-\frac{1}{3}\sqrt{5 - 2\sqrt{\frac{10}{7}}}$ & $\frac{322+13\sqrt{70}}{900}$ \\
\hline
$0$ & $\frac{128}{225}$ \\
\hline
$\frac{1}{3}\sqrt{5 - 2\sqrt{\frac{10}{7}}}$ & $\frac{322+13\sqrt{70}}{900}$ \\
\hline
$\frac{1}{3}\sqrt{5 + 2\sqrt{\frac{10}{7}}}$ & $\frac{322-13\sqrt{70}}{900}$ \\
\hline
\label{intro_table}
\end{tabular}
\end{center}
\begin{lstlisting}
from math import sqrt
points = np.array([- sqrt(5 + 2 * sqrt(10. / 7)) / 3,
                   - sqrt(5 - 2 * sqrt(10. / 7)) / 3,
                   0,
                   sqrt(5 - 2 * sqrt(10. / 7)) / 3,
                   sqrt(5 + 2 * sqrt(10. / 7)) / 3])
weights = np.array([(322 - 13 * sqrt(70)) / 900,
                    (322 + 13 * sqrt(70)) / 900,
                    128. / 225,
                    (322 + 13 * sqrt(70)) / 900,
                    (322 - 13 * sqrt(70)) / 900])
\end{lstlisting}

We now calculate the integral

\begin{lstlisting}
integral = (b - a)/2 * np.inner(weights, g(points))
\end{lstlisting}

\begin{problem}

Write a function that accepts an array of points, an array of weights, and limits of integration and returns the integral.

\end{problem}

Now, how do we find the weights and the points?
There are many publications that will give lists of points for various weight functions.
We will demonstrate how to find such a list using the Golub-Welsch algorithm.
This algorithm builds a tri-diagonal matrix.
The points are then the eigenvalues.
The weights are the length of $\left[a, b\right]$ times the first coordinate of each eigenvector squared.
We note that finding eigenvalues for a tridiagonal matrix is a well conditioned, relatively painless problem.
Using a good eigenvalue solver gives the Golub-Welsch algorithm a complexity of $O(n^2)$.

There is some interesting theory required to understand this algorithm.
We only provide a brief sketch here and point the reader to the paper at \url{http://gubner.ece.wisc.edu/gaussquad.pdf} for further details.
We mentioned that the choice of the weight function corresponds to a class of orthogonal polyomials.
An important fact about orthogonal polynomials is that any set of orthogonal polynomials $\{p_i\}_{i=1}^N$ satisfies a three term recurrence relation
\[
p_i(x) = (a_ix + b_i)p_{i-1}(x) - c_ip_{i-2}(x)
\]
where $p_{-1}(x) = 0$ and $p_0(x) = 1$.
The coefficients $\{a_i, b_i, c_i\}$ have been calculated for several classes of orthogonal polynomials, and may be determined for an arbitrary class using the procedure found in ``Calculation of Gauss Quadrature Rules'' by Golub and Welsch.
Using these coefficients we may create a tri-diagonal matrix

\[
J = \begin{bmatrix}

\alpha_1 & \beta_1 & 0 & 0 & ... & 0 \\
\beta_1 & \alpha_2 & \beta_2 & 0 & ... & 0 \\
0 & \beta_2 & \alpha_3 & \beta_3 & ... & 0 \\
\vdots & & & & & \vdots \\
\vdots & & & & & \vdots \\
0 & ... & & & & \beta_{N-1} \\
0 & ... & & & \beta_{N-1} & \alpha_N

\end{bmatrix}
\]

Where $\alpha_i = \frac{-b_i}{a_i}$ and $\beta_i = (\frac{c_{i+1}}{a_ia_{i+1}})^{\frac{1}{2}}$.
This matrix is called the Jacobi matrix.
Derivation of the Jacobi matrix can also be found in the Golub-Welsch paper.
The eigenvalues of this matrix give us the points $x_i$ and the first componenent (or first entry) of the corresponding eigenvectors give the weight.

\begin{problem}
Write a function that will accept three arrays representing the coefficients $\{a_i, b_i, c_i\}$ from the recurrence relation above.
Return the Jacobi matrix.
\end{problem}

\begin{problem}
The coefficients of the Legendre polynomials (which correspond to the weight function $W(x) = 1$ on $[-1,1]$ are given by
\begin{align*}
a_i &= \frac{2i - 1}{i} \\
b_i &= 0 \\
c_i &= \frac{i-1}{i}
\end{align*}

Using the Jacobi matrix function you just wrote, find the points $x_i$ and weights $w_i$ that correspond to this weight function.
Do they match the ones given in the first part of this lab?

\end{problem}

%Another exercise could be Chebyshev polynomials.  See wikipedia for the weight function.

There do exist other techniques for finding the weights and points for a given weighting function.
This is, in fact, not even the fastest method.
In general practice, we may use \li{scipy.integrate} to calculate integrals.
\li{scipy.integrate.quadrature} offers a reasonably fast Gaussian quadrature implementation.

